Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-16 arxiv/diff_cov.tex
final edits
minor grammar fixes

The previous section assumes that all tasks have the same sample size and feature vectors.
In this section, we study how different sample sizes and different features impact hard parameter sharing.
%The different covariates case differs from the same covariates case in two aspects.
%First, different tasks may have different sample sizes. In extreme scenarios, one task may have much less labeled data compared to another task.
The setting where features differ across tasks is often called ``covariate shift''.

Unlike the previous section, we no longer have an optimal solution to $f(A, B)$.
This is because $f(A, B)$ is in general non-convex.
Instead, our result implies sharp bias-variance tradeoffs for any \emph{local minimizer} of $f(A, B)$.
We focus on the two-task case to better understand the impact of having different sample sizes and different covariates.
Let $n_1, n_2$ denote task one  and two's sample size, respectively.
Suppose that
\begin{align*}
	X^{(1)} &= Z^{(1)}(\Sigma^{(1)})^{1/2} \in \real^{n_1 \times p}, \text{ and} \\
	X^{(2)} &= Z^{(2)}(\Sigma^{(2)})^{1/2} \in \real^{n_2 \times p},
\end{align*}
where the entries of $Z^{(1)}$ and $ Z^{(2)}$ are drawn independently from a one dimensional distribution with zero mean, unit variance, and constant $\varphi$-th moment for a fixed $\varphi > 4$. The matrices $\Sigma^{(1)}\in \R^{p\times p}$ and $\Sigma^{(2)}\in \R^{p\times p}$ denote the population covariance matrices of task 1 and task 2, respectively.

@@

\paragraph{How does hard parameter sharing scale with sample sizes and covariate shift $M$?}
One can see that the variance limit depends intricately on both tasks' samples sizes and covariate shift.
Next, we illustrate how varying them impact the prediction loss.
\begin{example}[Sample size ratio]\label{ex_sample_ratio}
	We first consider the impact of varying sample sizes.
	Consider the random-effects model from Section \ref{sec_same}, with both tasks having an isotropic population covariance matrix.

@@

%Recall that  $U\Lambda V^\top$ is the singular value decomposition of $M$, %$\Lambda$ consists of the singular values of $M$, $V$ is an orthonormal matrix,
%It is not hard to verify that $(n_1 + n_2)\hat{\Sigma}^{-1}= (\Sigma^{(2)})^{-1/2} V W^{-1}V^\top (\Sigma^{(2)})^{-1/2}$ and .
Let $\mu=p^{-1}\sum_{i} \delta_{\sigma_i}$ denote the empirical spectral distribution of $W$, where the $\sigma_i$'s are the eigenvalues of $W$ and $\delta_{\sigma_i}$ is the point mass measure at $\sigma_i$. Then it is easy to see that the Stieltjes transform of $\mu$ is equal to
 \[ m_{\mu}(z) \define \frac{1}{p}\sum_{i=1}^p \frac{1}{\sigma_i - z}= p^{-1}\tr\left[(W-z\id)^{-1}\right]. \]
The above matrix $(W - z\id)^{-1}$ is known as $W$'s resolvent or Green's function.
%When $p$ goes to infinity, it is well-known that $m_{\mu}(z)$ converges to a fixed distribution governed by a set of self-consistent equations.
%These self-consistent equations give the asymptotic limit of the trace of $\hat{\Sigma}^{-1}$.
%The above approach applies when $\Sigma^{(2)}$ is isotropic in Theorem \ref{thm_main_RMT}.
%Since our goal is to show the limit of $\tr\left[ (Y - z\id)^{-1}V^{\top}{\Sigma^{(2)}}^{-1}V \right]$ as shown in equation \eqref{eigen2extra}, we study the  $(Y-z\id)^{-1}$.
%Compared to the Stieltjes transform, the resolvent also applies to random matrices.
We prove the convergence of $W$'s resolvent using the so-called ``local law'' with a sharp convergence rate \cite{isotropic,erdos2017dynamical,Anisotropic}. The full details will be given in Section \ref{appendix RMT}.
%Recent developments in the random matrix literature have shown the convergence of the resolvent matrix using the so-called ``local laws'' or ``deterministic equivalents'' (cf. \cite{Hachem2007deterministic,DS18}).

>>>>

The previous section assumes that all tasks have the same sample size and feature vectors.
In this section, we study how having different sample sizes and different covariance matrices impact hard parameter sharing.
%The different covariates case differs from the same covariates case in two aspects.
%First, different tasks may have different sample sizes. In extreme scenarios, one task may have much less labeled data compared to another task.
The setting where covariates differ across tasks is often called ``covariate shift''.

Unlike the previous section, we can no longer characterize the global minimum of $f(A, B)$.
This is because $f(A, B)$ is in general non-convex.
Instead, our result implies sharp bias-variance tradeoffs for any \emph{local minimizer} of $f(A, B)$.
We focus on the two-task case to better understand the impact of having different sample sizes and different covariates.
Let $n_1, n_2$ denote task one  and two's sample size, respectively.
Suppose
\begin{align*}
	X^{(1)} &= Z^{(1)}(\Sigma^{(1)})^{1/2} \in \real^{n_1 \times p}, \text{ and} \\
	X^{(2)} &= Z^{(2)}(\Sigma^{(2)})^{1/2} \in \real^{n_2 \times p},
\end{align*}
where the entries of $Z^{(1)}$ and $ Z^{(2)}$ are drawn independently from a one dimensional distribution with zero mean, unit variance, and constant $\varphi$-th moment for a fixed $\varphi > 4$. $\Sigma^{(1)}\in \R^{p\times p}$ and $\Sigma^{(2)}\in \R^{p\times p}$ denote the population covariance matrices of task 1 and task 2, respectively.

@@

\paragraph{How does hard parameter sharing scale with sample sizes and covariate shift $M$?}
One can see that the variance limit depends intricately on both tasks' samples sizes and covariate shift.
Next, we illustrate how varying them impact the prediction loss.
\begin{example}[Sample size ratio]\label{ex_sample_ratio}
	We first consider the impact of varying sample sizes.
	Consider the random-effect model from Section \ref{sec_same}, with both tasks having an isotropic population covariance matrix.

@@

%Recall that  $U\Lambda V^\top$ is the singular value decomposition of $M$, %$\Lambda$ consists of the singular values of $M$, $V$ is an orthonormal matrix,
%It is not hard to verify that $(n_1 + n_2)\hat{\Sigma}^{-1}= (\Sigma^{(2)})^{-1/2} V W^{-1}V^\top (\Sigma^{(2)})^{-1/2}$ and .
Let $\mu=p^{-1}\sum_{i} \delta_{\sigma_i}$ denote the empirical spectral distribution of $W$, where the $\sigma_i$'s are the eigenvalues of $W$ and $\delta_{\sigma_i}$ is the point mass measure at $\sigma_i$. Then it is easy to see that the Stieltjes transform of $\mu$ is equal to
 \[ m_{\mu}(z) \define \frac{1}{p}\sum_{i=1}^p \frac{1}{\sigma_i - z}= p^{-1}\tr\left[(W-z\id)^{-1}\right]. \]
The above matrix $(W - z\id)^{-1}$ is known as $W$'s resolvent or Green's function.
%When $p$ goes to infinity, it is well-known that $m_{\mu}(z)$ converges to a fixed distribution governed by a set of self-consistent equations.
%These self-consistent equations give the asymptotic limit of the trace of $\hat{\Sigma}^{-1}$.
%The above approach applies when $\Sigma^{(2)}$ is isotropic in Theorem \ref{thm_main_RMT}.
%Since our goal is to show the limit of $\tr\left[ (Y - z\id)^{-1}V^{\top}{\Sigma^{(2)}}^{-1}V \right]$ as shown in equation \eqref{eigen2extra}, we study the  $(Y-z\id)^{-1}$.
%Compared to the Stieltjes transform, the resolvent also applies to random matrices.
We prove the convergence of $W$'s resolvent using the so-called ``local law'' with a sharp convergence rate \cite{isotropic,erdos2017dynamical,Anisotropic}. The complete proof is provided in Section \ref{appendix RMT}.
%Recent developments in the random matrix literature have shown the convergence of the resolvent matrix using the so-called ``local laws'' or ``deterministic equivalents'' (cf. \cite{Hachem2007deterministic,DS18}).

### 2020-10-16 arxiv/setup.tex
minor grammar fixes

\textbf{Proof techniques.}
There are two main ideas in our analysis. The proof of our first result uses a geometric intuition that hard parameter sharing finds a ``rank-$r$'' approximation of the datasets.
We carefully keep track of the concentration error between the global minimizer of $f(A, B)$ and its population version (cf. equation \eqref{eq_A_star}).
The proof of our second result is significantly more involved because of different sample sizes and covariate shifts. Using recently developed techniques from random matrix theory, we show that the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts converges to a deterministic diagonal matrix asymptotically (cf. Theorem \ref{thm_main_RMT}). % Moreover, we can obtain a sharp bound on the concentration error.
% to obtain a sharp estimate on the..., which is commonly referred to as the \emph{local law}.
%\HZ{add several sentences on the technical insight}
One limitation of our analysis is that in Example \ref{ex_sample_ratio}, there is an error term that can result in vacuous bounds for very small $n_1$ (cf. equation \eqref{cor_MTL_error}).
We believe our result has provided important initial insights and it is an interesting question to tighten our result.

@@

There is a large body of classical and recent works on multi-task learning.
We focus our discussion on theoretical works, and refer interested readers to several excellent surveys for general references \cite{PY09,ZY17,V20}.
The early work of \citet{B00,BS03,M06} studied multi-task learning from a theoretical perspective, often using uniform convergence or Rademacher complexity based techniques.
An influential paper by \citet{BBCK10} provides uniform convergence bounds that combine multiple datasets in certain settings.
One limitation of uniform convergence based techniques is that the results often assume that all  tasks have the same sample size, see e.g. \citet{B00,MPR16}.
Moreover, these techniques do not apply to the high-dimensional setting, because the results usually require a sample size at least $p \log p$.

Our proof techniques use the so-called local law of random matrices \cite{erdos2017dynamical}, which is a recent development in the random matrix theory literature.
In the single task case, \citet{isotropic} first proved such a local law for sample covariance matrices with isotropic covariance.
\citet{Anisotropic} later extended this result to arbitrary covariance setting.
%On the other hand, one may derive the asymptotic result in Theorem \ref{thm_main_RMT} with error $\oo(1)$ using the free addition of two independent random matrices in  theory .
These techniques provide almost sharp convergence rates to the asymptotic limit compared to other techniques such as free probability \cite{nica2006lectures}.
To the best of our knowledge, we are not aware of any previous results in the multi-task case, even for two tasks (with arbitrary covarite shifts).

The problem we study here is also related to high-dimensional prediction in transfer learning \cite{li2020transfer,bastani2020predicting} and distributed learning \cite{dobriban2018high}.
For example, \citet{li2020transfer} provides minimax optimal rates for predicting a target regression task given multiple sparse regression tasks.
One closely related work is \citet{WZR20}, which studied hard parameter sharing for two linear regression tasks.
However, their results only apply to sample size regimes at least logarithmic factors of dimension.
%\citet{WZR20} (and an earlier work by \citet{KD12}) observed that the shared layer size $r$ in hard parameter sharing plays a critical role of regularization.
%Linear models in multi-task learning have been studied in various settings, including online learning \cite{CCG10,DCSP18}, sparse regression \cite{LPTV09,LPVT11}, and representation learning \cite{BHKL19}.

@@

\textbf{Organizations.}
The rest of this paper is organized as follows.
In Section \ref{sec_same}, we present the bias-variance decomposition for hard parameter sharing.
In Section \ref{sec_diff}, we present our technical results that describe how varying sample sizes and covariate shifts impact hard parameter sharing using random matrix theory.
In Sections \ref{sec_simulation} and \ref{sec_text}, we validate our theory in simulations and real world classification tasks.
In Section \ref{sec_conclude}, we summarize our work and discuss future work.
Section \ref{app_proof_error_same_cov}, \ref{appendix RMT}, and \ref{app_iso_cov} present proofs of our results.

>>>>

\textbf{Proof techniques.}
There are two main ideas in our analysis. The proof of our first result uses a geometric intuition that hard parameter sharing finds a ``rank-$r$'' approximation of the datasets.
We carefully keep track of the concentration error between the global minimizer of $f(A, B)$ and its population version (cf. equation \eqref{eq_A_star}).
The proof of our second result is significantly more involved because of different sample sizes and covariate shifts. We show that the inverse of the sum of two sample covariance matrices with arbitrary covariate shifts converges to a deterministic diagonal matrix asymptotically (cf. Theorem \ref{thm_main_RMT}).
We use recently developed techniques from random matrix theory to show a sharp convergence rate.
% to obtain a sharp estimate on the..., which is commonly referred to as the \emph{local law}.
%\HZ{add several sentences on the technical insight}
One limitation of our analysis is that in Example \ref{ex_sample_ratio}, there is an error term that can result in vacuous bounds for very small $n_1$ (cf. equation \eqref{cor_MTL_error}).
We believe our result has provided significant initial insights, and it is an interesting question to tighten our result.

@@

There is a large body of classical and recent works on multi-task learning.
We focus our discussion on theoretical results and refer interested readers to several excellent surveys for general references \cite{PY09,ZY17,V20}.
The early work of \citet{B00,BS03,M06} studied multi-task learning from a theoretical perspective, often using uniform convergence or Rademacher complexity based techniques.
An influential paper by \citet{BBCK10} provides uniform convergence bounds that combine multiple datasets in certain settings.
One limitation of uniform convergence based techniques is that the results often assume that all  tasks have the same sample size, see e.g. \citet{B00,MPR16}.
Moreover, these techniques do not apply to the high-dimensional setting because the results usually require a sample size of at least $p \log p$.

Our proof techniques use the so-called local law of random matrices \cite{erdos2017dynamical}, a recent development in the random matrix theory literature.
In the single-task case, \citet{isotropic} first proved such a local law for sample covariance matrices with isotropic covariance.
\citet{Anisotropic} later extended this result to arbitrary covariance matrices.
%On the other hand, one may derive the asymptotic result in Theorem \ref{thm_main_RMT} with error $\oo(1)$ using the free addition of two independent random matrices in  theory .
These techniques provide almost sharp convergence rates to the asymptotic limit compared to other methods such as free probability \cite{nica2006lectures}.
To the best of our knowledge, we are not aware of any previous results in the multi-task case, even for two tasks (with arbitrary covariate shifts).

The problem we study here is also related to high-dimensional prediction in transfer learning \cite{li2020transfer,bastani2020predicting} and distributed learning \cite{dobriban2018high}.
For example, \citet{li2020transfer} provide minimax-optimal rates to predict a target regression task given multiple sparse regression tasks.
One closely related work is \citet{WZR20}, which studied hard parameter sharing for two linear regression tasks.
However, their results only apply to sample size regimes at least logarithmic factors of dimension.
%\citet{WZR20} (and an earlier work by \citet{KD12}) observed that the shared layer size $r$ in hard parameter sharing plays a critical role of regularization.
%Linear models in multi-task learning have been studied in various settings, including online learning \cite{CCG10,DCSP18}, sparse regression \cite{LPTV09,LPVT11}, and representation learning \cite{BHKL19}.

@@

\textbf{Organizations.}
The rest of this paper is organized as follows.
In Section \ref{sec_same}, we present the bias-variance decomposition for hard parameter sharing.
In Section \ref{sec_diff}, we describe how varying sample sizes and covariate shifts impact hard parameter sharing using random matrix theory.
In Sections \ref{sec_simulation}, we validate our results in simulations.
In Section \ref{sec_conclude}, we summarize our work and discuss future work.
Section \ref{sec_text} describes our study on text classification tasks.
Section \ref{app_proof_error_same_cov}, \ref{appendix RMT}, and \ref{app_iso_cov} present proofs of our results.
